Practical Optional Types for Clojure

### 2015-07-06 overview.tex
wrap up section 2

@@

@@

If a specific constructor, method, or field cannot be found based on the
static types of its arguments, a type error is thrown.

@@

The conditional guards from dereferencing \clj{nil}, and as before
removing it is a static type error.

@@

A \emph{multimethod} is a kind of extensible function. It combines
a \emph{dispatch function} with one or more \emph{methods}.
When invoked, the arguments are first supplied to the dispatch function, yielding
a \emph{dispatch value}. A method is then chosen
based on the dispatch value---this method is applied to the arguments
to finally return a value for the entire expression.

@@

The dispatch function \clj{class}---associated at multimethod creation with \clj{defmulti}---dictates 
whether the \clj{String} or \clj{File} method is chosen---both installed via \clj{defmethod}.

@@

\begin{minted}{clojure}
  (isa? String Object) ;=> true
  (isa? Object String) ;=> false
  ; keywords are interned symbols with a : prefix
  (isa? :a :a) ;=> true
  (isa? :a :b) ;=> false
\end{minted}

@@

Dispatching on a non-class value is common, as well as specifying
a \clj{:default} method for when no methods are matched.

\begin{minted}{clojure}
(ann hi [Kw -> Str])
(defmulti hi identity)
(defmethod hi :en [_] "hello")
(defmethod hi :fr [_] "bonjour")
(defmethod hi :default [_] "um...")

(str (hi :en) " " (hi :fr)) ;=> "hello bonjour"
(hi :bocce)   ;=> "um..."
\end{minted}

The \clj{hi} multimethod dispatches on its argument via the identity function.

@@

Immutable hash-maps with keyword keys play a major role in Clojure programming.

@@

breakfast ; :- (HMap :mandatory {:en Str :fr Str}
          ;          :complete? true)

@@

HMap types model the most common usages of keyword maps.
The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
and \clj{:fr} have string values (expressed as a \clj{:mandatory} entry) and that 
every other key is absent (by \clj{:complete?} being \clj{true}).

@@

Futhermore, no special attention is needed for multimethod dispatch---the primitive branching
mechanism is still the humble \clj{if} conditional. Only a small number of rules are needed
to encode the \clj{isa?}-based dispatch, themselves made of small, simple pieces.
In practice, this means all conditional-based control flow understood by occurrence typing
also extends to the context of multimethod dispatch, and vice-versa.

@@

\begin{minted}{clojure}
(ann desserts [Order -> Int])
(defmulti desserts "Number of desserts per order."
          :Meal)  ; dispatch on :Meal entry
; map destructuring backwards, d is :desserts entry
(defmethod desserts :lunch [{d :desserts}] d)
(defmethod desserts :dinner [{d :desserts}] d)
(defmethod desserts :combo [{m1 :meal1 m2 :meal2}]
  (+ (desserts m1) (desserts m2)))

@@

The \clj{:combo} method is verified to only structurally recur
on \clj{Order}s. This is achieved because we learn the argument
must be at least of type
\clj{'{:Meal :combo}}
since
\clj{(isa? (:Meal o) :combo)}
must be true. This eliminates possibility of \clj{:lunch} and \clj{:dinner}
orders, deducing
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
as the argument type, which can safely structurally recur.

\paragraph{Arbitrary dispatch}
An equally valid dispatch mechanism for \clj{desserts}
would be on the \clj{class} of the \clj{:desserts} key.
We have already seen dispatch on \clj{class} and on keywords
in isolation---occurrence typing automatically understands
control flow that combines its simple building blocks.

\begin{minted}{clojure}
(ann desserts' [Order -> Int])
(defmulti desserts' 
          (fn [o :- Order] (class (:desserts o))))
(defmethod desserts' Long [{d :desserts}] d)
(defmethod desserts' nil [{m1 :meal1 m2 :meal2}]
  (+ (desserts' m1) (desserts' m2)))

(desserts' {:Meal :combo 
            :meal1 {:Meal :lunch :desserts 1}
            :meal2 {:Meal :dinner :desserts 2}})
;=> 3
\end{minted}

@@

Combined with the knowledge that it is also an
\clj{Order}, we can eliminate the possibility of
a \clj{:combo} meal, and infer the type
\begin{minted}{clojure}
(U '{:Meal ':lunch :desserts Int}
   '{:Meal ':dinner :desserts Int})
\end{minted}
Looking up \clj{:desserts} on this type returns an \clj{Int}.

@@

Since lookups on missing keys return \clj{nil}, there are
now two possibilities:
\begin{itemize}
  \item either \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
  \item \clj{o} is missing a \clj{:desserts} entry.
\end{itemize}
In a type, we learn \clj{o} is at least
\begin{minted}{clojure}
(U '{:desserts nil}
   ; :absent-keys, a set of known absent entries
   (HMap :absent-keys #{:desserts}))
\end{minted}

This is enough to infer \clj{o} is a \clj{:combo} meal, thus verifying both
structural recursions.

>>>>

Typed Clojure guarantees that well-typed code cannot
dereference the null-pointer.
This is especially important for Clojure programs---\clj{nil}
is treated like any other distinct datum in Clojure,  and its status as
one of only two false values means it is a common choice to indicate
``nothing'' or ``failure''.

@@

The following Typed Clojure program constructs a new \clj{File}
instance and calls the \clj{getParent} method on the result.

@@

Typed Clojure helps the Clojure compiler avoid expensive reflective 
calls,
however if a specific constructor, method, or field cannot be found based on the
static types of its arguments, a type error is thrown.

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.

@@

A \emph{multimethod} is a kind of extensible function. It combines
a \emph{dispatch function} with one or more \emph{methods}.

\paragraph{Value-based dispatch}
Here is a simple multimethod.

\begin{minted}{clojure}
(ann hi [Kw -> Str]) ; interface type for `hi`
(defmulti hi identity) ; dispatch function `identity`
(defmethod hi :en [_] "hello") ; method for `:en`
(defmethod hi :fr [_] "bonjour") ; method for `:fr`
(defmethod hi :default [_] "um...") ; default method
\end{minted}

When invoked, the arguments are first supplied to the dispatch function---\clj{identity}---yielding
a \emph{dispatch value}. A method is then chosen
based on the dispatch value---this method is applied to the arguments
to finally return a value for the entire expression.

\begin{minted}{clojure}
(map hi [:en :fr :bocce]) ; map over keyword vector 
;=> ("hello" "bonjour" "um...")
\end{minted}

For example, 
\clj{(hi :en)} evaluates to \clj{"hello"}---it executes
the \clj{:en} method
because \clj{(= (identity :en) :en)} is true
and \clj{(= (identity :en) :fr)} is false.

This is only part of the story for multimethod dispatch.

\paragraph{Class-based dispatch}
For class values, multimethods choose methods based on subclassing
relationships.

@@

The dispatch function \clj{class}
%---associated at multimethod creation with \clj{defmulti}---
dictates 
whether the \clj{String} or \clj{File} method is chosen.
%---both installed via \clj{defmethod}

@@

\begin{minted}{clojure}
(isa? (identity :en) :en) ;=> true
(isa? (identity :en) :fr) ;=> false
(isa? (class "STAINS/JELLY") String) ;=> true
(isa? (class "STAINS/JELLY") Object) ;=> true
(isa? (class (new File "JELLY")) String) ;=> false
\end{minted}

@@

@@

Immutable hash-maps with keyword keys play a major role in Clojure programming
as the primary way to structure data.

@@

@@

HMap types model the most common usages of keyword maps.

\begin{minted}{clojure}
breakfast ; :- (HMap :mandatory {:en Str :fr Str}
          ;          :complete? true)
\end{minted}

The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
and \clj{:fr} have string values (expressed as a \clj{:mandatory} entry) and that 
every other key is absent (by \clj{:complete?} being \clj{true}).

@@

Futhermore, no special attention is needed for multimethod dispatch---the primitive branching
mechanism is still the humble \clj{if} conditional. Only a small number of rules are needed
to encode the \clj{isa?}-based dispatch, themselves made of small, simple pieces.
In practice, this means all conditional-based control flow analysis
also extends to the context of multimethod dispatch, and vice-versa.

@@

\begin{minted}{clojure}
(ann desserts [Order -> Int])
(defmulti desserts "Total desserts per order."
          :Meal)  ; dispatch on :Meal entry
; map destructuring backwards, d is :desserts entry
(defmethod desserts :lunch [{d :desserts}] d)
(defmethod desserts :dinner [{d :desserts}] d)
(defmethod desserts :combo [{m1 :meal1 m2 :meal2}]
  (+ (desserts m1) (desserts m2)))

@@

The \clj{:combo} method is verified to only structurally recur
on \clj{Order}s. This is achieved because we learn the argument--- \clj{o} ---must % em dash needs space here, looks weird
be of type
\clj{'{:Meal :combo}}
since
\clj{(isa? (:Meal o) :combo)}
must be true. Combining \clj{'{:Meal :combo}}
with the fact that \clj{o} is an \clj{Order}
eliminates possibility of \clj{:lunch} and \clj{:dinner}
orders, simplifying \clj{o} to
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
which can safely structurally recur.

\paragraph{Nested dispatch}
An equally valid dispatch mechanism for \clj{desserts}
would be on the \clj{class} of the \clj{:desserts} key.
We have already seen dispatch on \clj{class} and on keywords
in isolation---occurrence typing automatically understands
control flow that combines its simple building blocks.

\begin{minted}{clojure}
(ann desserts' [Order -> Int])
(defmulti desserts' 
          (fn [o :- Order] (class (:desserts o))))
(defmethod desserts' Long [{d :desserts :as o}] 
  ; o :- (U '{:Meal ':lunch :desserts Int}
  ;         '{:Meal ':dinner :desserts Int})
  d)
(defmethod desserts' nil [{m1 :meal1 m2 :meal2 :as o}]
  ; o :- '{:Meal ':combo :meal1 Order :meal2 Order}
  (+ (desserts' m1) (desserts' m2)))
\end{minted}
%
%(desserts' {:Meal :combo 
%            :meal1 {:Meal :lunch :desserts 1}
%            :meal2 {:Meal :dinner :desserts 2}})
%;=> 3

@@

Knowing \clj{o} is also an
\clj{Order}, we can eliminate the possibility of
a \clj{:combo} meal, and infer the type
\begin{minted}{clojure}
(U '{:Meal ':lunch :desserts Int}
   '{:Meal ':dinner :desserts Int})
\end{minted}
Looking up \clj{:desserts} on this type returns an \clj{Int}.

@@

Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
  \item \clj{o} is missing a \clj{:desserts} entry.
\end{itemize}
Equivalently, we learn \clj{o} is at least of type
\begin{minted}{clojure}
(U '{:desserts nil}
   ; :absent-keys, a set of known absent entries
   (HMap :absent-keys #{:desserts}))
\end{minted}
This intersected with \clj{o} being an \clj{Order} simplifies to a \clj{:combo} meal, 
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
thus verifying both structural recursions.

\paragraph{From multiple to arbitrary dispatch}
Multimethod dispatch goes even further, supporting
multiple dispatch via vectors.

Fully specifying multimethod dispatch is beyond the scope of this paper,
but the same intuition applies---adding support for multiple dispatch
automatically allows arbitrary combinations and nestings
of it and previous simple dispatch rules.
